\documentclass[11pt, a4paper]{article}
\usepackage{amsmath, amssymb}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{parskip}
\usepackage{titlesec}
\usepackage{xcolor}

\geometry{margin=3cm}
\titleformat{\section}{\large\bfseries}{}{0em}{}[\titlerule]
\titleformat{\subsection}{\normalsize\bfseries}{}{0em}{}

\hypersetup{colorlinks=true, linkcolor=black, urlcolor=blue!60!black}

\title{\textbf{Growing Metal Lace}\\
  \large From a Fractal Grammar to Evolutionary Geometry\\
  on Apple Silicon}
\author{Carlo Perassi}
\date{2026}

\begin{document}
\maketitle
\thispagestyle{empty}

\begin{abstract}
We describe SierpSphere, a project that starts with a compact grammar
for fractal signed-distance functions, adds a genetic algorithm targeting
real-world metal 3D-printing constraints, and accelerates the inner loop
with Apple Metal via PyTorch MPS.
The result is an automated search over a space of printable, aesthetically
complex fractal objects — each one expressible as a single line of symbolic
notation — running at interactive speed on a laptop.
\end{abstract}

% ── 1 ─────────────────────────────────────────────────────────────────────────
\section{Motivation}

The Sierpi\'nski gasket has been rendered millions of times.
What is less common is asking: \emph{what other objects live near it in the
space of self-similar Boolean-SDF grammars?}
And, once we can enumerate that neighbourhood, \emph{which of those objects
could actually exist as a centimetre-scale metal artefact?}

SierpSphere began as a weekend renderer and grew into an end-to-end pipeline
spanning grammar design, real-time browser visualisation, evolutionary search,
and GPU acceleration — all within a single repository, reproducible via Podman
or a single shell script.

% ── 2 ─────────────────────────────────────────────────────────────────────────
\section{The Grammar}

\subsection{Signed Distance Functions as a generative substrate}

An SDF $f : \mathbb{R}^3 \to \mathbb{R}$ implicitly defines a surface as
$\{x \mid f(x) = 0\}$.
Smooth Boolean operations — smooth-union, smooth-subtract, smooth-intersect
— compose SDFs without artifacts, and their blending parameter $k$ controls
whether the join is sharp or organic.

Polyhedral symmetry groups ($T_d$, $O_h$, $I_h$) provide a natural set of
axes along which to place child primitives.
At each iteration step the grammar places scaled copies of a primitive
(sphere, cube, or octahedron) along those axes and combines them with
the accumulated SDF via one of the three Boolean operations.

\subsection{A compact notation}

A grammar is a symmetry group, a seed primitive, and an ordered list of steps.
The full state needed to reconstruct any evolved object fits on one line:

\[
  \Gamma = \mathsf{G}[\mathrm{P_0}] : s_1 \cdot s_2 \cdots s_n
\]

For example, the classic Sierpi\'nski sphere is
$\mathsf{T_d}[\mathrm{S}]:\ominus_s^{0.5}{}_{0.02}\cdot\oplus_s^{0.5}{}_{0.01}\cdot\ominus_s^{0.5}{}_{0.005}$.
A POSIX-safe filename slug encoding (replacing special characters with
letters) makes this notation directly usable in gallery filenames:
\texttt{Td.S.Ns4Ps2Ns1}.

% ── 3 ─────────────────────────────────────────────────────────────────────────
\section{The Fitness Landscape}

\subsection{Beauty as a measurable quantity}

We approximate visual complexity with four metrics computed on the extracted
trimesh:

\begin{itemize}
  \item \textbf{Fractal dimension} — box-counting on mesh vertices, rewarded
        when the slope $d_f \in [2.3, 2.7]$ (Gaussian centred at 2.5).
  \item \textbf{Curvature variance} — discrete mean curvature measure;
        high variance implies rich surface detail.
  \item \textbf{Genus} — topological handles; a high-genus mesh has many
        through-holes, a characteristic of Sierpi\'nski-like lace.
  \item \textbf{Normalised surface-to-volume ratio} — rewards forms that
        fill space efficiently without being solid blobs.
\end{itemize}

\subsection{Manufacturability as a hard constraint}

Direct Metal Laser Sintering (DMLS) and Selective Laser Melting (SLM)
impose strict geometric requirements.
Three hard gates eliminate any candidate that fails them outright:

\begin{enumerate}
  \item \textbf{No islands} — the mesh must be a single connected component.
  \item \textbf{No enclosed voids} — sealed internal cavities trap metal
        powder and cannot be removed post-print.
  \item \textbf{Minimum wall thickness $\geq 0.3\,\mathrm{mm}$} — estimated
        via ray-casting on the surface at the target scale of $80\,\mathrm{mm}$.
\end{enumerate}

Soft manufacturing metrics — overhang fraction (support volume ratio),
thermal mass uniformity across octants, drain openings — contribute to the
weighted fitness score $f \in [0,1]$.

\subsection{The full fitness function}

Fifteen sub-scores are combined as a weighted sum.
Aesthetic metrics account for 53\% of the total weight;
manufacturing metrics for 47\%.
The two groups are deliberately balanced: an object that is beautiful but
unprintable scores zero, and a printable object that is visually trivial
scores poorly.

% ── 4 ─────────────────────────────────────────────────────────────────────────
\section{The Genetic Algorithm}

A population of $N=24$ grammars evolves over 50 epochs.
Each epoch:

\begin{enumerate}
  \item \textbf{Evaluate} — extract the mesh via marching cubes at resolution
        $28^3$ and compute the fitness.
  \item \textbf{Select} — tournament selection ($k=4$) with elitism (top 3
        carried unchanged).
  \item \textbf{Reproduce} — single-point crossover on the iteration-step
        list (50\% probability) followed by per-field Gaussian mutation
        (rate 0.25 per step).
\end{enumerate}

Mutation operators act on numeric parameters (scale, distance, smoothness)
with multiplicative Gaussian jitter, and on categorical fields
(operation type, primitive) with discrete swap.
Steps can also be added or removed, allowing the grammar length to evolve.

% ── 5 ─────────────────────────────────────────────────────────────────────────
\section{Metal Acceleration}

\subsection{The bottleneck}

An SDF grammar with $n$ operations evaluated on a $40^3$ grid requires
$40^3 \times n \approx 10^6$ floating-point evaluations per individual.
With a population of 24 and two CPU workers (inside a Podman Linux VM on
macOS), epoch time reached 2 hours — too slow for iterative experimentation.

\subsection{PyTorch MPS}

Apple Silicon exposes a unified memory GPU via Metal Performance Shaders.
PyTorch's MPS backend maps tensor operations onto Metal compute kernels
transparently.

The SDF evaluation is inherently data-parallel: for a grid of $N^3$ points,
every point is independent.
We reshape the $N^3$ grid into a flat $(N^3, 3)$ tensor and evaluate all
primitives and Boolean operations as batched tensor arithmetic — no Python
loop over voxels.

\begin{verbatim}
pts = torch.stack([gx, gy, gz], dim=1)   # (N³, 3) on MPS
sdf = sd_sphere(pts, center, r)
for op in grammar_ops:
    child = sd_primitive(pts, op)
    sdf   = smooth_subtract(sdf, child, k)
grid = sdf.reshape(N, N, N).cpu().numpy()
\end{verbatim}

The marching cubes step (scikit-image, CPU) and trimesh fitness metrics
remain on the CPU; only the SDF sampling moves to the GPU.
This is sufficient because SDF sampling dominates the per-individual cost.

\subsection{Observed speedup}

On an Apple M-series chip with MPS, the per-individual evaluation time
at resolution $28^3$ drops from $\sim 5\,\mathrm{min}$ (CPU, Podman VM)
to $\sim 3\,\mathrm{s}$ (Metal, native), a speedup of roughly $100\times$.
Epoch time for a population of 24 falls from hours to minutes.

% ── 6 ─────────────────────────────────────────────────────────────────────────
\section{System Architecture}

The project is structured as five independently usable components:

\begin{center}
\begin{tabular}{lll}
\toprule
Component & Technology & Purpose \\
\midrule
\texttt{engine/} & Python · Flask · trimesh & SDF eval · GLB export · REST API \\
\texttt{viewer/} & Three.js · GLSL · Nginx & Real-time MC viewer · HQ PBR snapshot \\
\texttt{evolver/} & Python · PyTorch MPS & GA · fitness · Metal eval \\
\texttt{glb-viewer/} & Three.js (standalone) & Material preview · turntable · 4K PNG \\
\texttt{tex/} & \LaTeX & Grammar notation · this document \\
\bottomrule
\end{tabular}
\end{center}

The Podman/Docker stack (\texttt{docker-compose.yml}) orchestrates the
containerised path; \texttt{run\_native.sh} provides a one-command native
macOS setup with automatic virtualenv creation.

% ── 7 ─────────────────────────────────────────────────────────────────────────
\section{Open Questions and Future Directions}

\begin{itemize}
  \item \textbf{Higher-resolution saves} — the winner of each epoch is
        saved at $64^3$; a post-evolution high-res pass at $256^3$ on Metal
        would produce print-ready geometry in seconds.
  \item \textbf{Multi-objective Pareto front} — rather than a scalar fitness,
        maintain a Pareto front of (beauty, printability) pairs and let the
        user navigate the trade-off interactively.
  \item \textbf{Grammar grammar} — the mutation operators themselves could
        be subject to evolution (hyper-heuristics), allowing the search to
        adapt its own exploration strategy.
  \item \textbf{Differentiable SDF} — replacing marching cubes with a
        differentiable renderer (e.g.\ sphere tracing with PyTorch autograd)
        would enable gradient-based optimisation of grammar parameters,
        potentially orders of magnitude faster than the GA for local refinement.
  \item \textbf{Physical simulation} — coupling the fitness function to a
        finite-element stress simulator would reward geometries that are not
        just printable but structurally sound under load.
\end{itemize}

% ── 8 (LinkedIn version) ──────────────────────────────────────────────────────
\section*{Short version (LinkedIn / general audience)}

\textit{What happens when you let a computer invent its own fractal sculptures
and grade them on whether they could be 3D printed in metal?}

I built SierpSphere: a system where a grammar defines fractal shapes as
sequences of sphere-carving operations, a genetic algorithm evolves thousands
of variants, and a fitness function scores each one on visual complexity
\emph{and} real manufacturing constraints — minimum wall thickness, no
floating islands, no trapped powder.

The twist: evaluating one shape required sampling a million points in 3D space.
On a standard CPU that took minutes per shape, making overnight runs impractical.
By rewriting the inner loop as a PyTorch tensor operation on Apple Metal (MPS),
the same evaluation runs in about 3 seconds — a 100$\times$ speedup without
touching the algorithm or the output quality.

The best shapes from each generation are saved as GLB files you can open
in any 3D viewer, and each one has a compact symbolic name (like a chemical
formula) that fully describes how to recreate it.

\medskip
Source: \url{https://github.com/carlok/sierpsphere}

\end{document}
